A simple and efficient algorithm for computing approximate Nash equilibria in sequential games with imcomplete information

### 2015-07-28 paper.tex
typo
multiple fixups

The sequence-form representation for two-person zero-sum games with
incomplete information was introduced in
\cite{koller1992complexity}, and the theory was further developed in
\cite{koller1994fast,von1996efficient,vonequilibrium} where it was
established that for such games, there exist sparse matrices
$A \in \mathbb{R}^{n_1 \times n_2}$, $E_1 \in \mathbb{R}^{l_1 \times
  n_1}$, $E_2 \in \mathbb{R}^{l_2 \times n_2}$, and vectors $e_1 \in
\mathbb{R}^{l_1}, e_2 \in \mathbb{R}^{l_2}$ such that $n_1$, $n_2$,
$l_1$, and $l_2$ are all linear in the size of the game tree (number
of states in the game) and such that Nash equilibria correspond to
pairs $(x, y)$ of \textit{realization plans} which solve the primal
LCP (Linear Convex Program)
\begin{equation}
  \begin{aligned}
     \text{minimize }\langle e_1, p\rangle, \hspace{.5em}\text{
       subject to: }(y,p) \in \mathbb{R}^{n_2} \times
     \mathbb{R}^{l_1},  y \ge 0, E_2y = e_2, -Ay + E_1^Tp \geq 0.
  \end{aligned}
  \label{eq:primal_pb}
\end{equation}

and the dual LCP
\begin{equation}
  \begin{aligned}
    \text{maximize}-\langle e_2, q\rangle,\hspace{.5em}\text{subject
      to: }(x,q) \in \mathbb{R}^{n_1} \times \mathbb{R}^{l_2},  x \ge
    0, E_1x = e_1, A^Tx + E_2^Tq \geq 0.
  \end{aligned}
  \label{eq:dual_pb}
\end{equation}
The vectors $p = (p_0, p_1, ..., p_{l_2 - 1}) \in \mathbb{R}^{l_2}$
and $q = (q_0, q_1, ..., q_{l_1 - 1}) \in \mathbb{R}^{l_1}$ are dual
variables. 
$A$ is the \textit{payoff matrix} and each $E_k$ is a matrix whose
entries are $-1$, $0$ or $1$, with exactly 1 entry per row which
equals $-1$ except for the first whose whose first entry is $1$ and all
the others are $0$. Each of the vectors $e_k$ is of the form $(1, 0, ..., 0)$.

@@

Developing on an alternative notion of approximate equilibria (see
Definition \ref{thm:cool_notion})
homologous to that presented in Definition \ref{thm:approx_nash}, we
device a primal-dual algorithm that (Algorithm \ref{Tab:algo}) for
computing approximate Nash equilibria in sequential two-person
zero-sum games with incomplete information and perfect recall, and
which satisfies the above crucial requirements (1)--(2). We also
prove (Theorem \ref{thm:pd}) that --in an ergodic sense-- the number of
iterations required by the algorithm to produce an approximation
equilibrium to a precision $\epsilon$ is $\mathcal{O}(1/\epsilon)$,
with explicit values for the constants involved this worst-case cost.
These contributions will be elaborated in section \ref{sec:gsp}.
%%  The source of approximation in our scheme
%% is that we explicitly control how close we are to the feasible set
%% $Q_1 \times Q_2$.

@@

The primal-dual algorithm first developed in \cite{chambolle2010}, was
proposed in \cite{chambolle2014ergodic} as a way of solving matrix
games on. It should be stressed that such matrix games on simplexes are
considerably simpler than the games considered here. Indeed, the
authors in \cite{chambolle2014ergodic} used the fact that computing the
orthogonal projection of a point onto a simplex can be done in linear
time as in \cite{duchi2008efficient}. In contrast, no such efficient
algorithm is known nor is likely to exist, for the polytopes $Q_k$
defined in \eqref{eq:polytope}.%% , the strategy profiles for players in
%% the games considered here.
That notwithstanding, such projections can still be done iteratively
using for example, the algorithm in proposition 4.2 of
\cite{combettes2010dualization} or the algorithms developed in
\cite{tran2015splitting}. Unfortunately, as with any nested iterative
scheme, one would have to solve this sub-problem with finer and finer
precision.

@@

\section{Methods}
\label{sec:gsp}
\subsection{The Generalized Saddle-point Problem (GSP) connection}
In the next theorem, we show that the LCPs \eqref{eq:primal_pb}
and \eqref{eq:dual_pb} can be
conveniently written as a \textit{Generalized Saddle-point Problem
  (GSP)} in the sense of \cite{he2013accelerating}.
\begin{theorem}
Define two proper closed convex functions
  \begin{eqnarray}
    \left.
    \begin{aligned}
      g_1: \mathbb{R}^{n_2} &\times \mathbb{R}^{l_1} \rightarrow
      (-\infty, +\infty], \hspace{1em} g_1(y, p) :=
        i_{y \ge 0} + \langle e_1,p\rangle\\
        g_2: \mathbb{R}^{n_1} &\times \mathbb{R}^{l_2} \rightarrow
        (-\infty, +\infty],\hspace{1em} g_2(x, q) :=
          i_{x \ge 0} + \langle e_2, q\rangle
    \end{aligned}
    \right\}
    \label{eq:things}
  \end{eqnarray}

@@

and, define the functions $\hat{\Psi}_1$, $\hat{\Psi}_2:
\mathbb{R}^{n_2} \times \mathbb{R}^{l_1} \times \mathbb{R}^{n_1}
\times \mathbb{R}^{l_2} \rightarrow (-\infty, +\infty]$ by
\begin{eqnarray}
  \left.
  \begin{aligned}
    \hat{\Psi}_1(y, p, x, q) &:=
    \Psi_1(y, p, x, q)+ g_1(y, p) \text{ if }y \ge 0, \hat{\Psi}_1(y,
    p, x, q) := \infty \text{ otherwise}\\
    \hat{\Psi}_2(y, p, x, q) &:= \Psi_2(y, p, x, q)+ g_2(x, q) \text{ if
    }x \ge 0, \hat{\Psi}_2(y,p, x, q) := \infty \text{ otherwise}.
  \end{aligned}
  \right\}
\end{eqnarray}

@@

\begin{algorithm}
\caption{Primal-dual algorithm for computing approximate Nash
  Equilbria in two-person zero-sum games with incomplete information
  and perfect recall}
\label{Tab:algo}
\begin{algorithmic}[1]
\Require $\epsilon > 0$; $(y^{(0)},p^{(0)},x^{(0)},q^{(0)})
\in \mathbb{R}^{n_2}
  \times \mathbb{R}^{l_1} \times \mathbb{R}^{n_1} \times
  \mathbb{R}^{l_2}$.
\Ensure A Nash $(\epsilon,0)$-equilibrium
$({y^*},{p^*},{x^*},{q^*}) \in S_1 \times S_2$ for
the GSP \eqref{eq:unconstrained_pb}.
\State  $\lambda \leftarrow 1/\|K\|$, ${v}_a^{(0)} \leftarrow 0$, $k
\leftarrow 0$
\While{$\|{v}_{a}^{(k)}\| \ge \epsilon$}
\State $y^{(k + 1)} \leftarrow (y^{(k)} - \lambda (A^Tx^{(k)} +
E_2^Tq^{(k)}))_+$, \hspace{.5em}$p^{(k+1)} \leftarrow p^{(k)} -
\lambda(e_1-E_1x^{(k)})$
\State $x^{(k + 1)} \leftarrow (x^{(k)} + \lambda (Ay^{(k+1)} -
E_1^Tp^{(k+1)}))_+$, \hspace{.5em}$\Delta x^{(k+1)} \leftarrow
x^{(k+1)}-x^{(k)}$
\State $\Delta q^{(k+1)} \leftarrow \lambda (E_2y -
e_2)$, \hspace{.5em}$q^{(k+1)} \leftarrow q^{(k)} + \Delta q^{(k+1)}$
\State $y^{(k+1)} \leftarrow y^{(k+1)} - \lambda (A^T\Delta x^{(k+1)}
+ E_2^T\Delta q^{(k+1)})$, \hspace{.5em}$\Delta y^{(k+1)} \leftarrow
y^{(k+1)}-y^{(k)}$
\State $p^{(k+1)} \leftarrow p^{(k+1)} + \lambda E_1\Delta x^{(k+1)}$,
\hspace{.5em} $\Delta p^{(k+1)} \leftarrow p^{(k+1)}-p^{(k)}$
\State ${v}_a^{(k+1)} \leftarrow \frac{1}{\lambda}\frac{1}{k+1}(y^{(k+1)} -
y^{(0)}, p^{(k+1)} - p^{(0)}, x^{(k+1)} - x^{(0)}, q^{(k+1)} - q^{(0)})$ 
\State $k \leftarrow k + 1$
\EndWhile
\end{algorithmic}
\end{algorithm}

\begin{theorem}[Ergodic $\mathcal{O}(1/\epsilon)$ convergence]
Let $d_0$ be the orthogonal distance between the starting point
$(y^{(0)},p^{(0)},x^{(0)},q^{(0)})$ of Algorithm \ref{Tab:algo} and the
set of equilibria for the GSP \eqref{eq:unconstrained_pb}.
Then given any $\epsilon > 0$, there exists an index
$k_0 \le \frac{2d_0\|K\|}{\epsilon}$ such that after $k_0$ iterations
the algorithm produces the following a quadruplet
$(y^{k_0},p^{k_0},x^{k_0},q^{k_0})$ and a vector $v_a^{k_0}$ such that
\begin{eqnarray}
\|v_a^{k_0}\| \le \epsilon \text{ and}\hspace{.5em} v_a^{k_0} \in
\partial[\hat{\Psi}_1(., ., x^{k_0}, q^{k_0}) +
  \hat{\Psi}_2(y^{k_0}, p^{k_0}, ., .)](y^{k_0},p^{k_0},x^{k_0},q^{k_0}).
\end{eqnarray}

Thus Algorithm \ref{Tab:algo} outputs an $(\epsilon,0)$-Nash
equilibrium for the GSP \eqref{eq:unconstrained_pb}
in at most $\frac{2d_0\|K\|}{\epsilon}$ iterations.
\end{theorem}

\begin{proof}
It is clear to see that the quadruplet $(\Psi_1, \Psi_2, g_1, g_2)$
satisfies assumptions B.1, B.2, B.3, B.5, and B.6 of
\cite{he2013accelerating} with $L_{xx} = L_{yy} = 0$ and $L_{xy} =
L_{yx} = \|K\|$. Now, one easily computes the proximal operator of
$g_j$ in closed-form as $\prox_{\lambda g_j}(a, b) \equiv ((a)_+,
b - \lambda e_j)$. With all these ingredients in place, Algorithm
\ref{Tab:algo} is then obtained from \cite[Algorithm
  T-BD]{he2013accelerating}, applied on the GSP
\eqref{eq:unconstrained_pb} with the choice of parameters: $\sigma = 1
\in (0, 1]$, $\sigma_x = \sigma_y = 0 \in [0, \sigma)$,
    $\lambda_{xy} := \frac{1}{\sigma L_{xy}}\sqrt{(\sigma^2 -
        \sigma_x^2)(\sigma^2 - \sigma_y^2)} = \sigma / \|K\| =
      1/\|K\|$, and $\lambda = \lambda_{xy} \in (0,
      \lambda_{xy}]$. Finally, note that $v_a^{(k+1)}$ as defined in
  line 8 of Algorithm \ref{Tab:algo} simply correponds to the the
  average difference in the iterates (this is a teloscopic sum),
  scaled by $\lambda$. The convergence result follows immediately from
  \cite[Theorem 4.2]{he2013accelerating}.
\end{proof}

>>>>

The sequence-form representation for two-person zero-sum games with
incomplete information was introduced in
\cite{koller1992complexity}, and the theory was further developed in
\cite{koller1994fast,von1996efficient,vonequilibrium} where it was
established that for such games, there exist sparse matrices
$A \in \mathbb{R}^{n_1 \times n_2}$, $E_1 \in \mathbb{R}^{l_1 \times
  n_1}$, $E_2 \in \mathbb{R}^{l_2 \times n_2}$, and vectors $e_1 \in
\mathbb{R}^{l_1}, e_2 \in \mathbb{R}^{l_2}$ such that $n_1$, $n_2$,
$l_1$, and $l_2$ are all linear in the size of the game tree (number
of states in the game) and such that Nash equilibria correspond to
pairs $(x, y)$ of \textit{realization plans} which solve the primal
LCP (Linear Convex Program)
\begin{equation}
  \begin{aligned}
    \underset{(y,p) \in \mathbb{R}^{n_2} \times
     \mathbb{R}^{l_1}}{\text{minimize }}\langle e_1,
    p\rangle \hspace{.5em}\text{
       subject to: } y \ge 0, E_2y = e_2, -Ay + E_1^Tp \geq 0.
  \end{aligned}
  \label{eq:primal_pb}
\end{equation}

and the dual LCP
\begin{equation}
  \begin{aligned}
    \underset{(x,q) \in \mathbb{R}^{n_1} \times
      \mathbb{R}^{l_2}}{\text{maximize }}-\langle e_2,
    q\rangle\hspace{.5em}\text{subject
      to: } x \ge
    0, E_1x = e_1, A^Tx + E_2^Tq \geq 0.
  \end{aligned}
  \label{eq:dual_pb}
\end{equation}
The vectors $p = (p_0, p_1, ..., p_{l_2 - 1}) \in \mathbb{R}^{l_2}$
and $q = (q_0, q_1, ..., q_{l_1 - 1}) \in \mathbb{R}^{l_1}$ are dual
variables. 
$A$ is the \textit{payoff matrix} and each $E_k$ is a matrix whose
entries are $-1$, $0$ or $1$, with exactly 1 entry per row which
equals $-1$ except for the first whose whose first entry is $1$ and all
the others are $0$. Each of the vectors $e_k$ is of the form $(1, 0, ..., 0)$.

@@

Developing on an alternative notion of approximate equilibria (see
Definition \ref{thm:cool_notion})
homologous to that presented in Definition \ref{thm:approx_nash}, we
device a primal-dual algorithm that (Algorithm \ref{Tab:algo}) for
computing approximate Nash equilibria in sequential two-person
zero-sum games with incomplete information and perfect recall, and
which satisfies the above crucial requirements (1)--(2). We also
prove (Theorem \ref{thm:pd}) that --in an ergodic / Ces\`ario sense-- the number of
iterations required by the algorithm to produce an approximation
equilibrium to a precision $\epsilon$ is $\mathcal{O}(1/\epsilon)$,
with explicit values for the constants involved this worst-case cost.
These contributions will be elaborated in section \ref{sec:gsp}.
%%  The source of approximation in our scheme
%% is that we explicitly control how close we are to the feasible set
%% $Q_1 \times Q_2$.

@@

The primal-dual algorithm first developed in \cite{chambolle2010}, was
proposed in \cite{chambolle2014ergodic} as a way of solving matrix
games on simplexes. It should be stressed that such matrix games on
simplexes are considerably simpler than the games considered
here. Indeed, the authors in \cite{chambolle2014ergodic} used the fact
that computing the orthogonal projection of a point onto a simplex can
be done in linear time as in \cite{duchi2008efficient}. In contrast,
no such efficient algorithm is known nor is likely to exist, for the
polytopes $Q_k$ defined in \eqref{eq:polytope}.%% , the strategy
%% profiles for players in %% the games considered here.
That notwithstanding, such projections can still be done iteratively
using for example, the algorithm in proposition 4.2 of
\cite{combettes2010dualization} or the algorithms developed in
\cite{tran2015splitting}. Unfortunately, as with any nested iterative
scheme, one would have to solve this sub-problem with finer and finer
precision.

@@

\section{Methods}
\label{sec:gsp}
\subsection{The Generalized Saddle-point Problem (GSP) connection}
In the next theorem, we show that the LCPs \eqref{eq:primal_pb}
and \eqref{eq:dual_pb} can be
conveniently written as a \textit{Generalized Saddle-point Problem
  (GSP)} in the sense of \cite{he2013accelerating}. The crux of idea
is to remove the linear constraints in the definitions of the strategy
polytopes $Q_k$, by modifying the payoff matrix to yield an equivalent
saddle-point problem.
\begin{theorem}
Define two proper closed convex functions
  \begin{eqnarray}
    \left.
    \begin{aligned}
      g_1: \mathbb{R}^{n_2} &\times \mathbb{R}^{l_1} \rightarrow
      (-\infty, +\infty], \hspace{1em} g_1(y, p) :=
        i_{y \ge 0} + \langle e_1,p\rangle\\
        g_2: \mathbb{R}^{n_1} &\times \mathbb{R}^{l_2} \rightarrow
        (-\infty, +\infty],\hspace{1em} g_2(x, q) :=
          i_{x \ge 0} + \langle e_2, q\rangle
    \end{aligned}
    \right\}
    \label{eq:things}
  \end{eqnarray}

@@

and define the functions $\hat{\Psi}_1$, $\hat{\Psi}_2:
\mathbb{R}^{n_2} \times \mathbb{R}^{l_1} \times \mathbb{R}^{n_1}
\times \mathbb{R}^{l_2} \rightarrow (-\infty, +\infty]$ by
\begin{eqnarray}
  \left.
  \begin{aligned}
    \hat{\Psi}_1(y, p, x, q) &:=
    \Psi_1(y, p, x, q)+ g_1(y, p) \text{ if }y \ge 0, \hat{\Psi}_1(y,
    p, x, q) := \infty \text{ otherwise}\\
    \hat{\Psi}_2(y, p, x, q) &:= \Psi_2(y, p, x, q)+ g_2(x, q) \text{ if
    }x \ge 0, \hat{\Psi}_2(y,p, x, q) := \infty \text{ otherwise}.
  \end{aligned}
  \right\}
\end{eqnarray}

@@

\begin{algorithm}
\caption{Primal-dual algorithm for computing approximate Nash
  Equilbria in two-person zero-sum games with incomplete information
  and perfect recall}
\label{Tab:algo}
\begin{algorithmic}[1]
\Require $\epsilon > 0$; $(y^{(0)},p^{(0)},x^{(0)},q^{(0)}) \in \mathbb{R}^{n_2}
  \times \mathbb{R}^{l_1} \times \mathbb{R}^{n_1} \times
  \mathbb{R}^{l_2}$.
\Ensure A Nash $(\epsilon,0)$-equilibrium
$({y^*},{p^*},{x^*},{q^*}) \in S_1 \times S_2$ for
the GSP \eqref{eq:unconstrained_pb}.
\State  $\lambda \leftarrow 1/\|K\|$, ${v}^{(0)} \leftarrow 0$, $k
\leftarrow 0$
\While{ $ k = 0$ or  $\frac{1}{k\lambda}\|v^{(k)}\| \ge \epsilon$}
\State $y^{(k + 1)} \leftarrow (y^{(k)} - \lambda (A^Tx^{(k)} +
E_2^Tq^{(k)}))_+$, \hspace{.5em}$p^{(k+1)} \leftarrow p^{(k)} -
\lambda(e_1-E_1x^{(k)})$
\State $x^{(k + 1)} \leftarrow (x^{(k)} + \lambda (Ay^{(k+1)} -
E_1^Tp^{(k+1)}))_+$, \hspace{.5em}$\Delta x^{(k+1)} \leftarrow
x^{(k+1)}-x^{(k)}$
\State $\Delta q^{(k+1)} \leftarrow \lambda (E_2y -
e_2)$, \hspace{.5em}$q^{(k+1)} \leftarrow q^{(k)} + \Delta q^{(k+1)}$
\State $y^{(k+1)} \leftarrow y^{(k+1)} - \lambda (A^T\Delta x^{(k+1)}
+ E_2^T\Delta q^{(k+1)})$, \hspace{.5em}$\Delta y^{(k+1)} \leftarrow
y^{(k+1)}-y^{(k)}$
\State $p^{(k+1)} \leftarrow p^{(k+1)} + \lambda E_1\Delta x^{(k+1)}$,
\hspace{.5em} $\Delta p^{(k+1)} \leftarrow p^{(k+1)}-p^{(k)}$
\State ${v}^{(k+1)} \leftarrow {v}^{(k)} + (\Delta
y^{(k+1)},\Delta p^{(k+1)},\Delta x^{(k+1)},\Delta q^{(k+1)})$ 
\State $k \leftarrow k + 1$
\EndWhile
\end{algorithmic}
\end{algorithm}

\begin{theorem}[Ergodic / Ces\`ario $\mathcal{O}(1/\epsilon)$ convergence]
Let $d_0$ be the orthogonal distance between the starting point
$(y^{(0)},p^{(0)},x^{(0)},q^{(0)})$ of Algorithm \ref{Tab:algo} and the
set of equilibria for the GSP \eqref{eq:unconstrained_pb}.
Then given any $\epsilon > 0$, there exists an index
$k_0 \le \frac{2d_0\|K\|}{\epsilon}$ such that after $k_0$ iterations
the algorithm produces the following a quadruplet
$(y^{k_0},p^{k_0},x^{k_0},q^{k_0})$ and a vector $v^{k_0}$ such that
\begin{eqnarray}
\|v_a^{k_0}\| \le \epsilon \text{ and}\hspace{.5em} v_a^{k_0} \in
\partial[\hat{\Psi}_1(., ., x^{k_0}, q^{k_0}) +
  \hat{\Psi}_2(y^{k_0}, p^{k_0}, ., .)](y^{k_0},p^{k_0},x^{k_0},q^{k_0}),
\end{eqnarray}
where $v_a^{(k_0)} := \frac{1}{k\lambda}v^{(k_0)}$.
Thus Algorithm \ref{Tab:algo} outputs an $(\epsilon,0)$-Nash
equilibrium for the GSP \eqref{eq:unconstrained_pb}
in at most $\frac{2d_0\|K\|}{\epsilon}$ iterations.
\end{theorem}

\begin{proof}
It is clear to see that the quadruplet $(\Psi_1, \Psi_2, g_1, g_2)$
satisfies assumptions B.1, B.2, B.3, B.5, and B.6 of
\cite{he2013accelerating} with $L_{xx} = L_{yy} = 0$ and $L_{xy} =
L_{yx} = \|K\|$. Now, one easily computes the proximal operator of
$g_j$ in closed-form as $\prox_{\lambda g_j}(a, b) \equiv ((a)_+,
b - \lambda e_j)$. With all these ingredients in place, Algorithm
\ref{Tab:algo} is then obtained from \cite[Algorithm
  T-BD]{he2013accelerating}, applied on the GSP
\eqref{eq:unconstrained_pb} with the choice of parameters: $\sigma = 1
\in (0, 1]$, $\sigma_x = \sigma_y = 0 \in [0, \sigma)$,
    $\lambda_{xy} := \frac{1}{\sigma L_{xy}}\sqrt{(\sigma^2 -
        \sigma_x^2)(\sigma^2 - \sigma_y^2)} = \sigma / \|K\| =
      1/\|K\|$, and $\lambda = \lambda_{xy} \in (0,
      \lambda_{xy}]$. The convergence result follows immediately from
  \cite[Theorem 4.2]{he2013accelerating}.
\end{proof}
